Kaggle-in-class Data Challenges Can Boost Student Learning

### 2020-04-29 paper-kaggle.Rmd
Some comments addressed

Kaggle [@kaggle] is a platform for predictive modelling and analytics competitions where participants compete to produce the best predictive model for a given data set. It is well-known for its competitions (e.g. @darkmatter), some of which come with rich monetary prizes (e.g. @heritage). There are also learning competitions [@kagglelearn], designed to help novices hone their data mining skills. Winners are typically expected to share their code, and occasionally new algorithms have emerged, for example, deep neural networks [@HintonDahi12] and XGBoost [@ChenXGBoost].

@@

A competition, like any other assessment method has its advantages and disadvantages. It brings the 'game' feeling, increases the interest level among students and motivates for higher performance (@Shindler09, p 105). But, it may have negative influence if constructed poorly. Among the negative influences are  increased stress and anxiety, induced by fearing a low ranking, failure or technology barriers. In addition, students may invest a disproportionate amount of time and effort into competition. Despite some received criticism, a properly set competition can benefit the students greatly. The competition should be relatively short in duration to avoid consuming undue energy. Students should be clear about the rules and the goal. They should be properly rewarded and most important, feel that they have reasonable chance to win or achieve high mark [@Shindler09].

@Shelley08 raised the need for more quantitative and statistical analysis of evidence in science education. This paper contributes to this call by offering statistical analysis of theffects on learning of classroom data competitions.

@@

Two data sets were compiled for the kaggle challenges: Melbourne property auction prices and spam classification. The Melbourne auction price data was collected by extracting information from real estate auction reports (pdf) collected between Feb 2, 2013 and Dec 17, 2016. The spam classification data was compiled by graduate students at Iowa State University as part of a data mining class, in 2009. Data was compiled by monitoring and extracting information from their emails by class members, over a period of a week, and manually tagging them as spam or ham.

@@

Computational Statistics and Data Mining (MAST90083) is designed for postgraduate level  students with math, statistics, information technology or actuarial backgrounds. It covers modelling both continuous (regression) and categorical (classification) response variables. The 63 students were randomized into one of two kaggle competitions, one focused on regression (R) and the other classification (C). This setup mimics randomized control trials, which is the 'gold standard, in experiment design (@Shelley09CH1, ch. 1).
Students built prediction models and made submissions individually for 16 days, and then were allowed to form groups to compete for another 7 days.

The reason for this strategy was first to motivate each of the students to think about modelling and be actively engaged in the competitions through individual submission. The lecturer allowed participants to create groups towards the end of the competition to illustrate the advantages of group work and mixing models. Another motivation for this stratagem was the university policy, requiring a strategy to assign students individually in group assignments.

@@

Statistical Thinking (ETC2420/5242), covers regression, but not classification, and has a mix of undergraduate and postgraduate students. Only the 34 postgraduate (5242) students were required to participate in the kaggle competition, and competed in the regression (R) challenge. This was run independently from the MAST90083 competition. The 145 undergraduate (2420) students were used as controls for examining performance of the postgraduate students. Although, it may be surprising at first glance, the undergraduate students provide a reasonable control for the graduate students. The class is taught to both cohorts simultaneously. The entry requirements to the Bachelors of Commerce at Monash is high, and these students have strong mathematics backgrounds. In the years prior to this experiment, the undergraduate scores on the final exam are indistinguishable from those of the graduate students. The competition ran for one month. Students formed their own teams of 2-4 members to compete. 
<!--
Several undergraduates also chose to compete individually. The material on regression methods, particularly the computational methods needed to successfully predict housing prices was new to both groups of students. 
-->
<!--
ETC3250, called Business Analytics, is an undergraduate course focusing on data mining. All students participated in a kaggle competition on the classification challenge. Because this group had no comparison group, it was difficult to assess performance. 
-->

@@

<!--
To examine the correlation between student’s engagement levels and the performances in the exam we plot the number of submission during the competition versus the performances in the exam, Figure \ref{fig:numsubmition}. Once again we exam the performances based on two normalizations. Once normalizing by the total possible marks for the relevant cluster of questions ($PTQ$) and once normalizing by total exam marks ($PTE$). For the students participated in the Melbourne Price competition is the cluster of regression questions. For the students participated in the Spam Classification competition is the cluster of questions about classification methods. 
-->

```{r numsubmition, echo = FALSE, fig.cap = "\\label{fig:numsubmition} Scatterplots of the exam performance (a-c) and competition performance (d-f) by number of prediction submissions, for the three student groups. The relationships with exam performance are weak. For the MAST90083 and ETC5242 regression competitions, a clear pattern is that predictions improved substantially with more submissions. (House price in ETC5242 were divided by 100,000, explaining the difference in magnitude of error between two competitions.)" , fig.width=6, fig.height=5}
etc5242 <- read_csv("data/competition_5242.csv")
scores_5242_comp <- scores_5242 %>% left_join(etc5242, by="id")
p1 <- ggplot(filter(UniMelbData, Proj_alloc == "classification"), aes(x=numsub, y=perf_class)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Classification performance") + 
  xlab("Num submissions") +
  xlim(c(1,5)) + ggtitle("a. MAST90083 C")
p2 <- ggplot(filter(UniMelbData, Proj_alloc == "regression"), aes(x=numsub, y=perf_reg)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Regression performance") + 
  scale_x_continuous("Num submissions", limits=c(1,13), breaks=seq(1, 13, 2)) + 
  ggtitle("b. MAST90083 R")
p3 <- ggplot(scores_5242_comp, aes(x=numsub, y=perf)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Regression performance") + 
  xlab("Num submissions") + ggtitle("c. ETC5242")
#grid.arrange(p1, p2, ncol=2)
p4 <- ggplot(filter(UniMelbData, Proj_alloc == "classification"), aes(x=numsub, y=PrivetScore)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Accuracy") + 
  xlab("Num submissions") +
  xlim(c(1,5)) + ggtitle("d. MAST90083 C")
p5 <- ggplot(filter(UniMelbData, Proj_alloc == "regression"), aes(x=numsub, y=PrivetScore)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  #ylab("Error") + 
  scale_x_continuous("Num submissions", limits=c(1,13), breaks=seq(1, 13, 2)) + 
  scale_y_continuous("Error ('0000s)", breaks=seq(110000, 170000, 10000), labels=seq(11,17,1)) + ggtitle("e. MAST90083 R")
  #xlim(c(1,12))
p6 <- ggplot(scores_5242_comp, aes(x=numsub, y=score)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Error") + 
  xlab("Num submissions") + ggtitle("f. ETC5242")
grid.arrange(p1, p2, p3, p4, p5, p6, ncol=3)

@@

1. the full set is not available to the students, to avoid cheating.
2. the data is not too easy, or too hard, to model so that there is some discriminatory power in the results.
3. data should be relatively clean, to the point where the instructor has tested that a model can be fitted.
4. contains some challenges, that make standard off-the-shelf modeling less successful, like different variable types that need processing or transforming, some outliers, large number of variables. 
5. if it is a classification challenge, it will work better with relatively balanced classes, because the overall accuracy is the easiest metric to use.

@@

Choosing the metric upon which to evaluate the model is another decision. Our advice is to keep it simple, so you, and the students, can understand the student scores. If you are running a regression challenge, then "Root Mean Squared Error (RMSE)". It is is a classification challenge, then "Categorization Accuracy", the percentage correct is reasonable.

@@

The competition needs to run without any intervention from the instructor. The instructor can monitor students progress: the number of submissions, student scores and even the uploaded data at any time. When the competition ends the "Leaderboard" page provides a list of students ordered by the final score. It also provides all the scores from all past submissions (under "Raw Data" on "Public Leaderboard").

@@

The primary finding, is that participating in a data challenge competition,  produces a statistically significant improvement in the learning of the topic, although the effect size is small.  Secondarily, anecdotally, the competitions enhanced interest and engagement in the course.

>>>>

Kaggle [@kaggle] is a platform for predictive modelling and analytics competitions where participants compete to produce the best predictive model for a given data set. It is well-known for its competitions (e.g. @darkmatter), some of which come with rich monetary prizes (e.g. @heritage). There are also learning competitions [@kagglelearn], designed to help novices hone their data mining skills. Winners are typically expected to share their code, and occasionally newly emerged algorithms are introduced to the broad community, for example, deep neural networks [@HintonDahi12] and XGBoost [@ChenXGBoost].

@@

A competition, like any other active learning method, that used for assessment, has its advantages and disadvantages. It brings the 'game' feeling, increases the interest level among students and motivates for higher performance (@Shindler09, p 105). However, it may have negative influence if constructed poorly. Among the negative influences are increased stress and anxiety, induced by fearing a low ranking, failure or technology barriers. In addition, students may invest a disproportionate amount of time and effort into competition. Despite some received criticism, a properly set competition can benefit the students greatly. The competition should be relatively short in duration to avoid consuming undue energy. Students should be clear about the rules and the goal. They should be properly rewarded and most important, feel that they have reasonable chance to win or achieve high mark [@Shindler09].

@Shelley08 raised the need for more quantitative and statistical analysis of evidence in science education. This paper contributes to this call by offering statistical analysis of the effects on learning of classroom data competitions.

@@

Two data sets were compiled for the kaggle challenges: Melbourne property auction prices and spam classification. The Melbourne auction price data was collected by extracting information from real estate auction reports (pdf) collected between Feb 2, 2013 and Dec 17, 2016. The spam classification data was compiled by graduate students at Iowa State University as part of a data mining class in 2009. Data was compiled by monitoring and extracting information from their emails by class members, over a period of a week, and manually tagging them as spam or ham.

@@

Computational Statistics and Data Mining (MAST90083) is designed for postgraduate level  students with math, statistics, information technology or actuarial backgrounds. It covers modelling both continuous (regression) and categorical (classification) response variables. The 63 students were randomized into one of two kaggle competitions, one focused on regression (R) and the other classification (C). This setup mimics randomized control trials, which is the gold standard, in experiment design (@Shelley09CH1, ch. 1).
Students built prediction models and made submissions individually for 16 days, and then were allowed to form groups to compete for another 7 days.

The reason for this strategy was first to motivate each of the students to think about modelling and be actively engaged in the competitions through individual submission. The lecturer allowed participants to create groups towards the end of the competition to illustrate the advantages of group work and ensemble models. Another reason for this stratagem was the university policy, requiring a strategy to assign students individually in group assignments. When the team members develop the model together, it is quite difficult to accurately assess the individual contribution of each student. The individual submissions helped to encourage each student to engage in the modelling process. In addition, it helped to assess the individual component of the final score for the competition.

@@

Statistical Thinking (ETC2420/5242), covers regression, but not classification, and has a mix of undergraduate and postgraduate students. Only the 34 postgraduate (5242) students were required to participate in the kaggle competition, and competed in the regression (R) challenge. This was run independently from the MAST90083 competition. The 145 undergraduate (2420) students were used as comparison for examining performance of the postgraduate students. Although, it may be surprising at first glance, the undergraduate students provide a reasonable comparison for the graduate students. The class is taught to both cohorts simultaneously. The entry requirements to the Bachelors of Commerce at Monash is high, and these students have strong mathematics backgrounds. In the years prior to this experiment, the undergraduate scores on the final exam are indistinguishable from those of the graduate students. The competition ran for one month. Students formed their own teams of 2-4 members to compete. 
<!--
Several undergraduates also chose to compete individually. The material on regression methods, particularly the computational methods needed to successfully predict housing prices was new to both groups of students. 
-->
<!--
ETC3250, called Business Analytics, is an undergraduate course focusing on data mining. All students participated in a kaggle competition on the classification challenge. Because this group had no comparison group, it was difficult to assess performance. 
-->

@@

<!--
To examine the correlation between students engagement levels and the performances in the exam we plot the number of submission during the competition versus the performances in the exam, Figure \ref{fig:numsubmition}. Once again we exam the performances based on two normalizations. Once normalizing by the total possible marks for the relevant cluster of questions ($PTQ$) and once normalizing by total exam marks ($PTE$). For the students participated in the Melbourne Price competition is the cluster of regression questions. For the students participated in the Spam Classification competition is the cluster of questions about classification methods. 
-->

```{r numsubmition, echo = FALSE, fig.cap = "\\label{fig:numsubmition} Scatterplots of the exam performance (a-c) and competition performance (d-f) by number of prediction submissions, for the three student groups. The relationships with exam performance are weak. For the MAST90083 and ETC5242 regression competitions, a clear pattern is that predictions improved substantially with more submissions. (House price in ETC5242 were divided by 100,000, explaining the difference in magnitude of error between two competitions.)" , fig.width=6, fig.height=5}

library(grid)

etc5242 <- read_csv("data/competition_5242.csv")
scores_5242_comp <- scores_5242 %>% left_join(etc5242, by="id")
p1 <- ggplot(filter(UniMelbData, Proj_alloc == "classification"), aes(x=numsub, y=perf_class)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Classification performance") + 
  xlab("Num submissions") +
  xlim(c(1,5)) + ggtitle("a. MAST90083",  subtitle = "Classification")
p2 <- ggplot(filter(UniMelbData, Proj_alloc == "regression"), aes(x=numsub, y=perf_reg)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Regression performance") + 
  scale_x_continuous("Num submissions", limits=c(1,13), breaks=seq(1, 13, 2)) + 
  ggtitle("b. MAST90083",  subtitle = "Regression")
p3 <- ggplot(scores_5242_comp, aes(x=numsub, y=perf)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Regression performance") + 
  xlab("Num submissions") + ggtitle("c. ETC5242",  subtitle = "Regression")
#grid.arrange(p1, p2, ncol=2)
p4 <- ggplot(filter(UniMelbData, Proj_alloc == "classification"), aes(x=numsub, y=PrivetScore)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Accuracy") + 
  xlab("Num submissions") +
  xlim(c(1,5)) + ggtitle("d. MAST90083",  subtitle = "Classification")
p5 <- ggplot(filter(UniMelbData, Proj_alloc == "regression"), aes(x=numsub, y=PrivetScore)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  #ylab("Error") + 
  scale_x_continuous("Num submissions", limits=c(1,13), breaks=seq(1, 13, 2)) + 
  scale_y_continuous("Error ('0000s)", breaks=seq(110000, 170000, 10000), labels=seq(11,17,1)) + ggtitle("e. MAST90083",  subtitle = "Regression")
  #xlim(c(1,12))
p6 <- ggplot(scores_5242_comp, aes(x=numsub, y=score)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Error") + 
  xlab("Num submissions") + ggtitle("f. ETC5242",  subtitle = "Regression")

#gtext <- grid::textGrob("MAST90083 R - Regression challenge, \nMAST90083 C - Classification challenge,\nETC5242 - Regression challenge", x =unit(0.3, "npc"), y = unit(0.7, "npc") , 
#                        just = "left") 
#grid.arrange(p1, p2, p3, p4, p5, p6, gtext, ncol=3)

grid.arrange(p1, p2, p3, p4, p5, p6,  ncol=3)

@@

1. the full set is not available to the students, to avoid plagiarism and use of unauthorized assistance.
2. the data is not too easy, or too hard, to model so that there is some discriminatory power in the results.
3. data should be relatively clean, to the point where the instructor has tested that a model can be fitted.
4. contains some challenges, that make standard off-the-shelf modeling less successful, like different variable types that need processing or transforming, some outliers, large number of variables. 
5. if it is a classification challenge, it will work better with relatively balanced classes, because the overall accuracy is the easiest metric to use.

@@

Choosing the metric upon which to evaluate the model is another decision. Our advice is to keep it simple, so you, and the students, can understand the student scores. If you are running a regression challenge, then the "Root Mean Squared Error (RMSE)" is a good choice. If it is a classification challenge, then "Categorization Accuracy", the percent of correct classifications, is reasonable.

@@

The competition needs to run without any intervention from the instructor. 
Exception is, of course, an academic discussion motivated by the competition between the teaching team and the students. For example, a discussion about different models, their advantages and limitations. 
The instructor can monitor students progress: the number of submissions, student scores and even the uploaded data at any time. When the competition ends the "Leaderboard" page provides a list of students ordered by the final score. It also provides all the scores from all past submissions (under "Raw Data" on "Public Leaderboard").

@@

The primary finding, is that participating in a data challenge competition,  produces a statistically significant improvement in the learning of the topic, although the effect size is small.  Secondarily, the competitions enhanced interest and engagement in the course.
